Automatically proving equivalence by type-safe reflection

### 2016-01-19 paper-JFP-DTP/content/2-basicIdeas.tex
corrections with Edwin advices on latex

\section{A simplified problem : type safe reflection for proving equalities with universally quantified natural numbers and additions}

We will first present the basic ideas of our technique on a simplified problem, in which we aim to deal with universally quantified natural numbers and additions.
For example, we would like to be able to automatically generate proofs of goals like $\forall x1\ x2\ x3:Nat,\ (x1 + l2) + (x3 + x4) = (x1 + (x2 + x3)) + x4$ [Example 1]. \\
For this smaller problem, we have decided to only work with the associativity of $+$ : $plusAssociative : \forall x1\ x2\ x3,\ (x1 + x2) + x3 = x1 + (x2 + x3)$ and with the fact that Z is a right neutral element for the $+$ operation : $plusZeroRightNeutral : \forall x, x + Z = x$. Note that $Z$ is also a left neutral element, but this property is not needed because we have this behaviour by reduction, as $+$ is defined recursively on its first argument. Of course, some datatypes can have more or less properties than these two, and this work will be extended in sections 3 and 4 where we will present a very general hierarchy of provers for multiple algebraic structures. \\
Thus, in definitive, in this section, we want to write a decision procedure, able to tell if two expressions composed of universally quantified natural numbers and additions of these numbers are equal, and to produce a proof of this equality if appropriate, where "equal" has the meaning "syntactically equal or equal thanks to the associativity and right neutral properties".

@@

When trying to prove this kind of equalities, the variables are abstracted, and they become part of the context. On the example 1 just above, after abstraction of the variables, the goal becomes simply $(x1 + x2) + (x3 + x4) = (x1 + (x2 + x3)) + x4$, which is something of the general form $x=y$.
The general idea --that will also apply for the more general problem detailed in the next sections-- will be to normalize both sides of the "potential equality" $x=y$, and afterwards to compare them using Leibniz syntactical equality.
The goal of the normalization is to compute a canonical representation for a number $x$, such that any other number provably equal to $x$ (by using the two available properties) will have the same canonical representation. For example, the normalisation might transform $x+((y+Z)+z)$ into $(x+y)+z$ if we decide that the normal form will be completely left associative, and that every addition between an element $a$ and zero should be simplified to $a$. It will then be possible to decide the equality by simply comparing the normalised left and right hand sides with a simple and strict syntactical equality. This is in fact what we do all the time when we have to decide if two things, written differently, are equal or not. When a human is given two mathematical polynomials and has to decide the equality of these functions, a technique that always works is to decide once and for all a canonical representation of polynomials, and to put both polynomials in this form. If the normalised forms are the same, then the two original polynomials are equal, otherwise they do not represent the same computation.\\
\\
In fact, such a normalisation function can't be written directly, because in the LHS and RHS of $x=y$, we potentially have variables which have been universally quantified. And the normalization function needs to do different treatments for a "variable natural number" (a number which has been universally quantified) and for the constant $Z$. This is not possible yet, because once the variables are abstracted, they are just ordinary values of type $Nat$, and there is no way to distinguish them. Indeed, this information only exists at the level of the ASTs representing the two terms, and we do not have a direct access to these syntax trees.
For this reason, we will work by reflection. In this little example, it means that we will define a datatype that will be used as an encoding of natural numbers, or more precisely, an encoding of natural numbers composed of "variable numbers", $Z$, and additions of these things. This datatype will let us inspect the internal structure of a number by pattern matching.
Previously, we were only able to pattern match a natural number against the constructors $Z$ and $S$, which wasn't what we needed.  With the first approximation of the data type $Expr$ presented in the Fig 6, we will be allowed to pattern match an encoding of number against the constructors $Plus$, $Var$ and $Zero$.

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
data Expr : ($\Gamma$ : Vect n Nat) $\rightarrow$ Type where
     Plus : {n:Nat} $\rightarrow$ {$\Gamma$:Vect n Nat} $\rightarrow$ 
            Expr $\Gamma$ $\rightarrow$ Expr $\Gamma$ $\rightarrow$ Expr $\Gamma$
     Var  : {n:Nat} $\rightarrow$ {$\Gamma$:Vect n Nat} $\rightarrow$ 
            (i : Fin n) $\rightarrow$ Expr $\Gamma$
     Zero : {n:Nat} $\rightarrow$ ($\Gamma$:Vect n Nat) $\rightarrow$ 
            Expr $\Gamma$ Z
\end{lstlisting}
\end{center}
\caption{First version of reflected natural numbers}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
e1 : (x, y, z : Nat) $\rightarrow$ 
           Expr [x, y, z]
e1 x y z = Plus (Plus (Var FZ) (Var (FS FZ))) 
                (Plus (Var FZ) (Var (FS (FS FZ))))
\end{lstlisting}
\end{center}
\caption{Reflected LHS of example 1 (with the first version of reflected natural numbers)}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
using (x : Nat, y : Nat, $\Gamma$ : Vect n Nat)
  data Expr : (Vect n Nat) $\rightarrow$ Nat $\rightarrow$ Type where
       Plus : Expr $\Gamma$ x $\rightarrow$ Expr $\Gamma$ y $\rightarrow$ 
              Expr $\Gamma$ (x + y)
       Var  : (i : Fin n) $\rightarrow$ Expr $\Gamma$ (index i $\Gamma$)
       Zero : Expr $\Gamma$ Z
\end{lstlisting}
\end{center}
\caption{Second version of reflected number with embedded denotation}
\figrule
\end{figure}

For an expression $ex\ :\ Expr\ \Gamma\ x$, we will say that "$ex$ denotes (or encodes) the number $x$ in the context $\Gamma$".
When an expression is a variable, the denoted number is simply the corresponding variable in the context, ie, $(index\ i\ \Gamma)$.
Also, the $Zero$ expression denotes the natural number $Z$.
Finally, if $ex$ is an expression encoding the number $x$, and $ey$ is an expression encoding the number $y$, then the expression $Plus\ ex\ ey$ denotes the concrete number $(x + y)$.

@@

We want to write the reduction function on a "correct by construction" way, which means that no additional proof should be required after the definition of the function. Thus, $reduce$ will produce the proof that the new Expr freshly produced has the same interpretation as the original Expr, and this will be made easier by the fact that the data type Expr is now indexed over the real --concrete-- number : a term of type $Expr\ \Gamma\ x$ is the encoding of the number $x$.
Thus, we can write the type of $reduce$ like this : \\
$reduce\ :\ Expr\ \Gamma\ x\ \rightarrow\ (x'\ **\ (Expr\ \Gamma\ x',\ x\ =\ x'))$ \\
The function $reduce$ produces a dependent pair : the new concrete number $x'$, and a pair made of an $Expr\ \Gamma\ x'$ which is the new encoded term indexed over the new concrete number we have just produced, and a proof that old and new --concrete-- numbers are equal.
Note that this function can't simply produce an $Expr\ \Gamma\ x$, because the concrete number on which the resulting expression will be indexed is not necessary syntactically equal to the original number since this equality can use the two available properties. Said differently, even if we can prove $x=x'$ (if the function is correctly defined), we do not have $x \equiv x'$.
And in fact, what really interests us in this function is precisely this proof of $x\ =\ x'$.
The reason is that when we try to automatically prove $x=y$, these proofs will be the crucial part for the construction of the desired proof. \\
\\
The tactic will work as follow.
We have an expression $ex$ encoding $x$, and an expression $ey$ encoding $y$. We will normalize $ex$, and this will give a new concrete number $x'$, a new expression $ex':Expr\ \Gamma\ x'$, and a proof of $x=x'$. We will do the same with $ey$, and we will get a new concrete number $y'$, an expression $ey':Expr\ \Gamma\ y'$, and a proof of $y=y'$. \\
It is now enough to simply compare $ex'$ and $ey'$ using a standard syntactical equality because these two expressions are now supposed to be in normal form :

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
eqExpr : (e : Expr $\Gamma$ x) $\rightarrow$ (e' : Expr $\Gamma$ y) $\rightarrow$ Maybe (e = e')
eqExpr (Plus x y) (Plus x' y') with (eqExpr x x', eqExpr y y')
  eqExpr (Plus x y) (Plus x y) | (Just Refl, Just Refl) = Just Refl
  eqExpr (Plus x y) (Plus x' y') | _ = Nothing
eqExpr (Var i) (Var j) with (decEq i j)
  eqExpr (Var i) (Var i) | (Yes Refl) = Just Refl
  eqExpr (Var i) (Var j) | _ = Nothing
eqExpr Zero Zero = Just Refl
eqExpr _ _ = Nothing
\end{lstlisting}
\end{center}
\caption{Syntactical equality between reflected terms}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
buildProof : {x : Nat} $\rightarrow$ {y : Nat} $\rightarrow$ Expr $\Gamma$ x' $\rightarrow$ Expr $\Gamma$ y' 
           $\rightarrow$ (x = x') $\rightarrow$ (y = y') $\rightarrow$ Maybe (x = y)
buildProof ex' ey' lp rp with (eqExpr ex' ey')
  buildProof ex' ex' lp rp | Just Refl = ?MbuildProof
  buildProof ex' ey' lp rp | Nothing = Nothing
\end{lstlisting}
\end{center}
\caption{Building the desired proof with the two proofs of equality}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
  MbuildProof = proof {
  intros; refine Just; rewrite sym p1; rewrite sym p2; exact Refl;
}  
\end{lstlisting}
\end{center}
\caption{buildProof metavariable}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
  testEq : Expr $\Gamma$ x $\rightarrow$ Expr $\Gamma$ y $\rightarrow$ Maybe (x = y)
  testEq ex ey = 
     let (x' ** (ex', px)) = reduce ex in 
     let (y' ** (ey', py)) = reduce ey in
        buildProof ex' ey' px py 
\end{lstlisting}
\end{center}
\caption{testEq}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
data LExpr : ($\Gamma$ : Vect n Nat) $\rightarrow$ Nat $\rightarrow$ Type where
     LPlus : LExpr $\Gamma$ x $\rightarrow$ (i : Fin n) 
             $\rightarrow$ LExpr $\Gamma$ (x + index i $\Gamma$)
     LZero : LExpr $\Gamma$ Z
\end{lstlisting}
\end{center}
\caption{Reflected left associative numbers}
\figrule
\end{figure}

@@

\end{lstlisting}
\end{center}
\caption{Production of the left associative form}
\figrule
\end{figure}
In the case of an addition $Plus\ ex\ ey$, the function $expr\_l$ does the job of normalisation recursively on $ex$ and on $ey$, and then it uses the sub-function $plusLExpr$ to normalise the addition of these two --already normalised-- terms. This sub-function $plusLExpr$ has two kind of simplifications to do. When the second argument is an $LZero$, it simply returns its first arguments along with the justification for this rewriting, which obviously uses $plusZeroRightNeutral$. However, when the second argument is an $LPlus\ e\ i$, it continues recursively by computing $plusLExpr\ ex\ e$, and it finally adds $i$ to it. That had for effect to move the parenthesis on the left, and this treatment is going to be justified during the proof of the meta variable $MplusLExpr$ by the use of $plusAssociative$.

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
MplusLExpr = proof {
  intros
  rewrite (sym (plusAssociative x1 x2 (index i $\Gamma$))); rewrite prfRec; 
  exact Refl;
}
\end{lstlisting}
\end{center}
\caption{Proof of the metavariable MplusLExpr}
\figrule
\end{figure}

It is really important to understand that the root of the automatic building of the desired proof of $x=y$ happens precisely in these nested use of $plusZeroRightNeutral$ and $plusAssociative$ that we've seen in the definition of $expr\_l$ and of its meta-variable $MplusLExpr$. These proofs replace the arithmetical proofs that we were doing previously by hand, as it was the case in section 1 with the lemma $adc\_lemma\_2$. \\
\\
Using this new datatype $LExpr$ has changed the representation of our encoded lists, so we need to convert back an $LExpr\ \Gamma\ x$ to an $Expr\ \Gamma\ x$. The function $l\_expr$ does this easy task.
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
l_expr : LExpr $\Gamma$ x $\rightarrow$ Expr $\Gamma$ x
l_expr LZero = Zero
l_expr (LPlus x i) = Plus (l_expr x) (Var i)
\end{lstlisting}
\end{center}
\caption{Going back from LExpr to Expr}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
  reduce : Expr $\Gamma$ x $\rightarrow$ (x' ** (Expr $\Gamma$ x', x = x'))
  reduce e = 
     let (x' ** (e', prf)) = expr_l e in
         (x' ** (l_expr e', prf))
\end{lstlisting}
\end{center}
\caption{Reduction function}
\figrule
\end{figure}

@@

\subsection{Usage of the "tactic"}

It's now time to see how to use this minimalist "tactic".
Let's define two expressions $e1$ and $e2$, respectively representing the numbers $((x + y) + (x + z))$ and $(x + ((y + x) + z))$ in the context $[x, y, z]$ of three abstracted variables.

@@

e2 : (x, y, z : Nat) 
     $\rightarrow$ Expr [x, y, z] (x + ((y + x) + z))
e2 x y z = Plus (Var FZ) 
                (Plus (Plus (Var (FS FZ)) 
                            (Var FZ)) 
                      (Var (FS (FS FZ))))
\end{lstlisting}
\end{center}
\caption{Two test expressions}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
e1_e2_testEq : (x, y, z : Nat) 
             $\rightarrow$ Maybe (((x + y) + (x + z)) = (x + ((y + x) + z)))
e1_e2_testEq x y z = testEq (e1 x y z) (e2 x y z)
\end{lstlisting}
\end{center}
\caption{Generating automatically the desired proof}
\figrule
\end{figure}

@@

\x => \y => \z => Just (replace (sym (replace (sym (replace 
(sym (plusAssociative x 0 y)) (replace (replace (sym 
(plusZeroRightNeutral x)) Refl) Refl))) (replace (sym 
(replace (sym (plusAssociative x 0 z)) (replace (replace 
(sym (plusZeroRightNeutral x)) Refl) Refl))) (replace (sym 
(plusAssociative (x+y) x z)) [...]
: (x : Nat) -> (y : Nat) -> (z : Nat) 
  -> Maybe ((x + y) + x + z 
            = x + (y + x) + z)
\end{lstlisting}
\end{center}
\caption{Obtained proof}
\figrule
\end{figure}

@@

\subsection{Construction of the reflected terms}

@@

We still want an automatic way of constructing these reflected terms because what we have so far is enough to demonstrate that our "prover" works, but that's not enough for being used intensively. We want to program an automatic way of going from the concrete values (of type $Nat$), to the reflected terms (of type $Expr$). The only way to do that is to inspect the abstract syntax tree of the concrete value.
By using Idris's reflection mechanism, we can tag a function with the keyword "\%reflection", which means that this function runs on syntax instead of values.

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting} [mathescape]
%reflection
total
reflectNat : {n:Nat} $\rightarrow$ ($\Gamma$ : Vect n Nat) $\rightarrow$ (x:Nat) 
             $\rightarrow$ (m ** ($\Gamma'$ : Vect m Nat ** (Expr ($\Gamma$ ++ $\Gamma'$) x)))
\end{lstlisting}
\end{center}
\caption{Type of the function for reflecting natural numbers}
\figrule
\end{figure}

@@

 \begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting} [mathescape]
reflectNat {n=n} G Z = (Z ** ([] ** (Zero {n=n+0} {G=G++[]})))
\end{lstlisting}
\end{center}
\caption{Reflecting natural numbers - pattern for zero}
\figrule
\end{figure}

@@

 \begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting} [mathescape]
reflectNat $\Gamma$ (x + y) =
     let (_ ** ($\Gamma'$ ** ex)) = (reflectNat $\Gamma$ x) in
     let (_ ** ($\Gamma''$ ** ey)) = (reflectNat ($\Gamma$ ++ $\Gamma'$) y) in
     let result = Plus (weaken $\Gamma''$ ex) ey in 
        (_ ** (($\Gamma'$ ++ $\Gamma''$) ** ?MreflectNat_1))
\end{lstlisting}
\end{center}
\caption{Reflecting natural numbers - pattern for a plus}
\figrule
\end{figure}

@@

 \begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting} [mathescape]
convertFin : (n:Nat) $\rightarrow$ (i:Fin n) $\rightarrow$ (x:Nat) $\rightarrow$ Fin (n+x)
\end{lstlisting}
\end{center}
\caption{Conversion of Fin}
\figrule
\end{figure}

@@

\end{lstlisting}
\end{center}
\caption{Reflecting natural numbers - pattern for a variable}
\figrule
\end{figure}

>>>>

\section{A Simplified Problem: Proving Equalities with Natural Numbers and Additions}

\label{sect:ideas}

We will first present the basic ideas of our technique on a simplified problem, in which we aim to deal with universally quantified natural numbers and additions.
For example, we would like to be able to automatically generate proofs of goals like $\forall x1\ x2\ x3:Nat,\ (x1 + l2) + (x3 + x4) = (x1 + (x2 + x3)) + x4$ [Example 1]. \\
For this smaller problem, we have decided to only work with the associativity of $+$ : $plusAssociative : \forall x1\ x2\ x3,\ (x1 + x2) + x3 = x1 + (x2 + x3)$ and with the fact that Z is a right neutral element for the $+$ operation : $plusZeroRightNeutral : \forall x, x + Z = x$. Note that $Z$ is also a left neutral element, but this property is not needed because we have this behaviour by reduction, as $+$ is defined recursively on its first argument. Of course, some datatypes can have more or less properties than these two, and this work will be extended in sections 3 and 4 where we will present a very general hierarchy of provers for multiple algebraic structures. \\
Thus, in definitive, in this section, we want to write a decision procedure, able to tell if two expressions composed of universally quantified natural numbers and additions of these numbers are equal, and to produce a proof of this equality if appropriate, where ``equal" has the meaning syntactically equal or equal thanks to the associativity and right neutral properties.

@@

When trying to prove this kind of equalities, the variables are abstracted, and they become part of the context. On the example 1 just above, after abstraction of the variables, the goal becomes simply $(x1 + x2) + (x3 + x4) = (x1 + (x2 + x3)) + x4$, which is something of the general form $x=y$.
The general idea --that will also apply for the more general problem detailed in the next sections-- will be to normalize both sides of the "potential equality" $x=y$, and afterwards to compare them using Leibniz syntactical equality.
The goal of the normalization is to compute a canonical representation for a number $x$, such that any other number provably equal to $x$ (by using the two available properties) will have the same canonical representation. For example, the normalisation might transform $x+((y+Z)+z)$ into $(x+y)+z$ if we decide that the normal form will be completely left associative, and that every addition between an element $a$ and zero should be simplified to $a$. It will then be possible to decide the equality by simply comparing the normalised left and right hand sides with a simple and strict syntactical equality. This is in fact what we do all the time when we have to decide if two things, written differently, are equal or not. When a human is given two mathematical polynomials and has to decide the equality of these functions, a technique that always works is to decide once and for all a canonical representation of polynomials, and to put both polynomials in this form. If the normalised forms are the same, then the two original polynomials are equal, otherwise they do not represent the same computation.\\
\\
In fact, such a normalisation function can't be written directly, because in the LHS and RHS of $x=y$, we potentially have variables which have been universally quantified. And the normalization function needs to do different treatments for a ``variable natural number" (a number which has been universally quantified) and for the constant $Z$. This is not possible yet, because once the variables are abstracted, they are just ordinary values of type $Nat$, and there is no way to distinguish them. Indeed, this information only exists at the level of the ASTs representing the two terms, and we do not have a direct access to these syntax trees.
For this reason, we will work by reflection. In this little example, it means that we will define a datatype that will be used as an encoding of natural numbers, or more precisely, an encoding of natural numbers composed of ``variable numbers", $Z$, and additions of these things. This datatype will let us inspect the internal structure of a number by pattern matching.
Previously, we were only able to pattern match a natural number against the constructors $Z$ and $S$, which wasn't what we needed.  With the first approximation of the data type $Expr$ presented in the figure~\ref{reflectedNaturalNumbers0}, we will be allowed to pattern match an encoding of number against the constructors $Plus$, $Var$ and $Zero$.

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
data Expr : ($\Gamma$ : Vect n Nat) $\rightarrow$ Type where
     Plus : {n:Nat} $\rightarrow$ {$\Gamma$:Vect n Nat} $\rightarrow$ 
            Expr $\Gamma$ $\rightarrow$ Expr $\Gamma$ $\rightarrow$ Expr $\Gamma$
     Var  : {n:Nat} $\rightarrow$ {$\Gamma$:Vect n Nat} $\rightarrow$ 
            (i : Fin n) $\rightarrow$ Expr $\Gamma$
     Zero : {n:Nat} $\rightarrow$ ($\Gamma$:Vect n Nat) $\rightarrow$ 
            Expr $\Gamma$ Z
\end{lstlisting}
\end{center}
\caption{First version of reflected natural numbers}
\label{reflectedNaturalNumbers0}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
e1 : (x, y, z : Nat) $\rightarrow$ 
           Expr [x, y, z]
e1 x y z = Plus (Plus (Var FZ) (Var (FS FZ))) 
                (Plus (Var FZ) (Var (FS (FS FZ))))
\end{lstlisting}
\end{center}
\caption{Reflected LHS with the first version of reflected numbers showed in \ref{reflectedNaturalNumbers0}}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
using (x : Nat, y : Nat, $\Gamma$ : Vect n Nat)
  data Expr : (Vect n Nat) $\rightarrow$ Nat $\rightarrow$ Type where
       Plus : Expr $\Gamma$ x $\rightarrow$ Expr $\Gamma$ y $\rightarrow$ 
              Expr $\Gamma$ (x + y)
       Var  : (i : Fin n) $\rightarrow$ Expr $\Gamma$ (index i $\Gamma$)
       Zero : Expr $\Gamma$ Z
\end{lstlisting}
\end{center}
\caption{Second version of reflected number with embedded denotation}
\label{reflectedNaturalNumbers}
\figrule
\end{figure}

For an expression $ex\ :\ Expr\ \Gamma\ x$, we will say that ``$ex$ denotes (or encodes) the number $x$ in the context $\Gamma$".
When an expression is a variable, the denoted number is simply the corresponding variable in the context, ie, $(index\ i\ \Gamma)$.
Also, the $Zero$ expression denotes the natural number $Z$.
Finally, if $ex$ is an expression encoding the number $x$, and $ey$ is an expression encoding the number $y$, then the expression $Plus\ ex\ ey$ denotes the concrete number $(x + y)$.

@@

We want to write the reduction function on a ``correct by construction" way, which means that no additional proof should be required after the definition of the function. Thus, $reduce$ will produce the proof that the new Expr freshly produced has the same interpretation as the original Expr, and this will be made easier by the fact that the data type Expr is now indexed over the real --concrete-- number : a term of type $Expr\ \Gamma\ x$ is the encoding of the number $x$.
Thus, we can write the type of $reduce$ like this : \\
$reduce\ :\ Expr\ \Gamma\ x\ \rightarrow\ (x'\ **\ (Expr\ \Gamma\ x',\ x\ =\ x'))$ \\
The function $reduce$ produces a dependent pair : the new concrete number $x'$, and a pair made of an $Expr\ \Gamma\ x'$ which is the new encoded term indexed over the new concrete number we have just produced, and a proof that old and new --concrete-- numbers are equal.
Note that this function can't simply produce an $Expr\ \Gamma\ x$, because the concrete number on which the resulting expression will be indexed is not necessary syntactically equal to the original number since this equality can use the two available properties. Said differently, even if we can prove $x=x'$ (if the function is correctly defined), we do not have $x \equiv x'$.
And in fact, what really interests us in this function is precisely this proof of $x\ =\ x'$.
The reason is that when we try to automatically prove $x=y$, these proofs will be the crucial part for the construction of the desired proof. \\
\\
The tactic will work as follow.
We have an expression $ex$ encoding $x$, and an expression $ey$ encoding $y$. We will normalize $ex$, and this will give a new concrete number $x'$, a new expression $ex':Expr\ \Gamma\ x'$, and a proof of $x=x'$. We will do the same with $ey$, and we will get a new concrete number $y'$, an expression $ey':Expr\ \Gamma\ y'$, and a proof of $y=y'$. \\
It is now enough to simply compare $ex'$ and $ey'$ using a standard syntactical equality because these two expressions are now supposed to be in normal form :

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
eqExpr : (e : Expr $\Gamma$ x) $\rightarrow$ (e' : Expr $\Gamma$ y) $\rightarrow$ Maybe (e = e')
eqExpr (Plus x y) (Plus x' y') with (eqExpr x x', eqExpr y y')
  eqExpr (Plus x y) (Plus x y) | (Just Refl, Just Refl) = Just Refl
  eqExpr (Plus x y) (Plus x' y') | _ = Nothing
eqExpr (Var i) (Var j) with (decEq i j)
  eqExpr (Var i) (Var i) | (Yes Refl) = Just Refl
  eqExpr (Var i) (Var j) | _ = Nothing
eqExpr Zero Zero = Just Refl
eqExpr _ _ = Nothing
\end{lstlisting}
\end{center}
\caption{Syntactical equality between reflected terms}
\label{syntacticalEq}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
buildProof : {x : Nat} $\rightarrow$ {y : Nat} $\rightarrow$ Expr $\Gamma$ x' $\rightarrow$ Expr $\Gamma$ y' 
           $\rightarrow$ (x = x') $\rightarrow$ (y = y') $\rightarrow$ Maybe (x = y)
buildProof ex' ey' lp rp with (eqExpr ex' ey')
  buildProof ex' ex' lp rp | Just Refl = ?MbuildProof
  buildProof ex' ey' lp rp | Nothing = Nothing
\end{lstlisting}
\end{center}
\caption{Building the desired proof with the two proofs of equality}
\label{buildProof}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
  MbuildProof = proof {
  intros; refine Just; rewrite sym p1; rewrite sym p2; exact Refl;
}  
\end{lstlisting}
\end{center}
\caption{buildProof metavariable}
\label{MbuildProof}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
  testEq : Expr $\Gamma$ x $\rightarrow$ Expr $\Gamma$ y $\rightarrow$ Maybe (x = y)
  testEq ex ey = 
     let (x' ** (ex', px)) = reduce ex in 
     let (y' ** (ey', py)) = reduce ey in
        buildProof ex' ey' px py 
\end{lstlisting}
\end{center}
\caption{testEq}
\label{testEq}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
data LExpr : ($\Gamma$ : Vect n Nat) $\rightarrow$ Nat $\rightarrow$ Type where
     LPlus : LExpr $\Gamma$ x $\rightarrow$ (i : Fin n) 
             $\rightarrow$ LExpr $\Gamma$ (x + index i $\Gamma$)
     LZero : LExpr $\Gamma$ Z
\end{lstlisting}
\end{center}
\caption{Reflected left associative numbers}
\label{LExpr}
\figrule
\end{figure}

@@

\end{lstlisting}
\end{center}
\caption{Production of the left associative form}
\label{expr_l}
\figrule
\end{figure}
In the case of an addition $Plus\ ex\ ey$, the function $expr\_l$ does the job of normalisation recursively on $ex$ and on $ey$, and then it uses the sub-function $plusLExpr$ to normalise the addition of these two --already normalised-- terms. This sub-function $plusLExpr$ has two kind of simplifications to do. When the second argument is an $LZero$, it simply returns its first arguments along with the justification for this rewriting, which obviously uses $plusZeroRightNeutral$. However, when the second argument is an $LPlus\ e\ i$, it continues recursively by computing $plusLExpr\ ex\ e$, and it finally adds $i$ to it. That had for effect to move the parenthesis on the left, and this treatment is going to be justified during the proof of the meta variable $MplusLExpr$ by the use of $plusAssociative$.

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
MplusLExpr = proof {
  intros
  rewrite (sym (plusAssociative x1 x2 (index i $\Gamma$))); rewrite prfRec; 
  exact Refl;
}
\end{lstlisting}
\end{center}
\caption{Proof of the metavariable MplusLExpr}
\label{MplusExpr}
\figrule
\end{figure}

It is really important to understand that the root of the automatic building of the desired proof of $x=y$ happens precisely in these nested use of $plusZeroRightNeutral$ and $plusAssociative$ that we've seen in the definition of $expr\_l$ and of its meta-variable $MplusLExpr$. These proofs replace the arithmetical proofs that we were doing previously by hand, as it was the case in section 1 with the lemma $adc\_lemma\_2$. \\
\\
Using this new datatype $LExpr$ has changed the representation of our encoded lists, so we need to convert back an $LExpr\ \Gamma\ x$ to an $Expr\ \Gamma\ x$. The function $l\_expr$ does this easy task.
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
l_expr : LExpr $\Gamma$ x $\rightarrow$ Expr $\Gamma$ x
l_expr LZero = Zero
l_expr (LPlus x i) = Plus (l_expr x) (Var i)
\end{lstlisting}
\end{center}
\caption{Going back from LExpr to Expr}
\label{l_expr}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
  reduce : Expr $\Gamma$ x $\rightarrow$ (x' ** (Expr $\Gamma$ x', x = x'))
  reduce e = 
     let (x' ** (e', prf)) = expr_l e in
         (x' ** (l_expr e', prf))
\end{lstlisting}
\end{center}
\caption{Reduction function}
\label{reduce}
\figrule
\end{figure}

@@

\subsection{Usage of the ``tactic"}

It's now time to see how to use this minimalist ``tactic".
Let's define two expressions $e1$ and $e2$, respectively representing the numbers $((x + y) + (x + z))$ and $(x + ((y + x) + z))$ in the context $[x, y, z]$ of three abstracted variables.

@@

e2 : (x, y, z : Nat) 
     $\rightarrow$ Expr [x, y, z] (x + ((y + x) + z))
e2 x y z = Plus (Var FZ) 
                (Plus (Plus (Var (FS FZ)) 
                            (Var FZ)) 
                      (Var (FS (FS FZ))))
\end{lstlisting}
\end{center}
\caption{Two test expressions}
\label{e1_e2}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
e1_e2_testEq : (x, y, z : Nat) 
             $\rightarrow$ Maybe (((x + y) + (x + z)) = (x + ((y + x) + z)))
e1_e2_testEq x y z = testEq (e1 x y z) (e2 x y z)
\end{lstlisting}
\end{center}
\caption{Generating automatically the desired proof}
\label{e1_e2_testEq}
\figrule
\end{figure}

@@

\x => \y => \z => Just (replace (sym (replace (sym (replace 
(sym (plusAssociative x 0 y)) (replace (replace (sym 
(plusZeroRightNeutral x)) Refl) Refl))) (replace (sym 
(replace (sym (plusAssociative x 0 z)) (replace (replace 
(sym (plusZeroRightNeutral x)) Refl) Refl))) (replace (sym 
(plusAssociative (x+y) x z)) [...]
: (x : Nat) -> (y : Nat) -> (z : Nat) 
  -> Maybe ((x + y) + x + z 
            = x + (y + x) + z)
\end{lstlisting}
\end{center}
\caption{Obtained proof}
\label{obtainedProof}
\figrule
\end{figure}

@@

\subsection{Construction of the reflected terms}
\label{sect:ReflectNat}

@@

We still want an automatic way of constructing these reflected terms because what we have so far is enough to demonstrate that our prover works, but that's not yet enough for being used intensively. We want to program an automatic way of going from the concrete values (of type $Nat$), to the reflected terms (of type $Expr$). The only way to do that is to inspect the abstract syntax tree of the concrete value.
By using Idris's reflection mechanism, we can tag a function with the keyword "\%reflection", which means that this function runs on syntax instead of values.

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting} [mathescape]
%reflection
total
reflectNat : {n:Nat} $\rightarrow$ ($\Gamma$ : Vect n Nat) $\rightarrow$ (x:Nat) 
             $\rightarrow$ (m ** ($\Gamma'$ : Vect m Nat ** (Expr ($\Gamma$ ++ $\Gamma'$) x)))
\end{lstlisting}
\end{center}
\caption{Type of the function for reflecting natural numbers}
\label{reflectNatType}
\figrule
\end{figure}

@@

 \begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting} [mathescape]
reflectNat {n=n} G Z = (Z ** ([] ** (Zero {n=n+0} {G=G++[]})))
\end{lstlisting}
\end{center}
\caption{Reflecting natural numbers - pattern for zero}
\label{reflectNat_pattern1}
\figrule
\end{figure}

@@

 \begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting} [mathescape]
reflectNat $\Gamma$ (x + y) =
     let (_ ** ($\Gamma'$ ** ex)) = (reflectNat $\Gamma$ x) in
     let (_ ** ($\Gamma''$ ** ey)) = (reflectNat ($\Gamma$ ++ $\Gamma'$) y) in
     let result = Plus (weaken $\Gamma''$ ex) ey in 
        (_ ** (($\Gamma'$ ++ $\Gamma''$) ** ?MreflectNat_1))
\end{lstlisting}
\end{center}
\caption{Reflecting natural numbers - pattern for a plus}
\label{reflectNat_pattern2}
\figrule
\end{figure}

@@

 \begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting} [mathescape]
convertFin : (n:Nat) $\rightarrow$ (i:Fin n) $\rightarrow$ (x:Nat) $\rightarrow$ Fin (n+x)
\end{lstlisting}
\end{center}
\caption{Conversion of a Fin}
\label{convertFin}
\figrule
\end{figure}

@@

\end{lstlisting}
\end{center}
\caption{Reflecting natural numbers - pattern for a variable}
\label{reflectNat_pattern3}
\figrule
\end{figure}
